Thermal reactionome of a temperate ant species

### 2014-05-15 doc/ApTranscriptome_manuscript.Rmd
Update to manuscript

While characterizing the reactionome is useful for description of the molecular pathways for thermal tolerance, a greater understanding of the process of evolution can be gained using a comparative approach [@harvey1991]. Specifically, comparisons of populations or species in contrasting environments can reveal the extent to which geographic differentiation and selection has changed the number of genes involved in thermal tolerance. To accomplish this, we compared the thermal reactionomes between the two species of *Aphaenogaster*. *A. picea*, the northern species, experiences less chronic and acute heat stress than *A. carolensis*. Thus, we predicted that *A. picea* would have more genes than *A. carolinensis* that are up-regulated to high temperatures, while *A. carolinensis* would have higher constitutive expression of these thermally-tolerant genes. Reciprocally, we expected *A. carolinensis* to have greater up-regulation of genes in response to cold temperatures, while *A. picea* would have greater constitutive expression of cold-tolerant genes. Finally, phenotypic studies in *Aphaenogaster* [@warren2013] and other ecotherms [@hoffmann2013] have found less variation in critical maximum than minimum temperatures. From this, we hypothesized that if the species had more similar maximum temperatures than minimum temperatures, the northern species *A. picea* would have a broader thermal breadth of gene expression than *A. carolinensis*.

@@

To gain greater insight on the genetic basis of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two species focusing on the thermally-responsive genes that had a significant species x temperature interaction. For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, *High*, *Low*, *Intermediate*, and *Bimodal*, as well as a fifth category *Not Responsive*. To focus on differences in gene expression between the two species, we performed GSEA for the set difference of each category.

@@

To test if genes that were responsive to cold temperatures in the southern *A. carolinensis* species or hot temperatures in the northern *A. picea* species had higher constitutive expression in the contrasting species, we performed non-parametric two-sided Wilcoxon tests comparing expression values near the optimum of 19.5° C.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
options(scipen = 1, digits = 3)
load("../ms.RData")
library(pander)
library(ggplot2)
```

The Trinity *de novo* transcriptome assembly included 126,172 genes with a total length of 100 million bp. Filtering to remove redundant or chimeric reads resulted in an assembly with 105,536 genes. The total length was 63 million bp with an N~50~ length of 895 bp and a mean gene size of 593 bp. Of the 105,536 filtered genes, 55,432 had hits to the NCBI-nr database. Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage. DeconSeq identified 5,675 genes as known contaminants from human, virus, bacterial or archael sources, leaving 99,861 clean transcripts. Of these, 47% have best hits to genes from the class Insecta [Supplementary Krona plot](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).

@@

We identified 22,399 genes with overall model fits of P < 0.05, retaining 10,597 after correcting for FDR. Of these transcripts, 9,156 significantly change expression with temperature including 6,908 that have a species x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the species ($\chi^2 = 170, P < 0.0001$; Table 1). For both species, the majority of responsive transcripts were in the *Low* category (Fig. 1). More transcripts had *Intermediate* expression in *A. carolinensis* than in *A. picea*, which had nearly twice as many bimodally-expressed transcripts  (Fig. 1, Table 1).

```{r expression_type_table, echo=FALSE, results='asis'}
pandoc.table(tt2, style="rmarkdown", split.tables = Inf, caption = "Number of thermally-responsive genes in each expression category for each species.")
```

@@

```{r expression_type_plot, echo=FALSE, fig.cap = "Mosaic plot that shows the proportion of thermally-responsive genes in each expression category for each species."}
mosaicplot(tt2, shade = FALSE, xlab = "ApVT", ylab = "AcNC", main = "", cex = 0.5)
```

As the molecular pathways involved in thermal response to cold and hot temperatures may differ, we performed gene set enrichment analysis on each expression category separately. Out of a total of 2,859 GO terms scored, we found 6 enriched (*P* < 0.01) for *High* transcripts, included categories implicated in stress response such as "gene expression" and "tissue regeneration" (Table S1). For *Low* genes, 60 GO terms were enriched including some implicated in stress response (e.g. detection of external stimulus; Table S1). For *Bimodal* genes, 11 GO terms were enriched including 10 "regulation" terms (e.g. regulation of primary metabolic process) indicating that these are general genes involved in activating responses to change (Table S1). Finally, for *Intermediate* genes 6 GO terms were enriched, all some type of metabolic or biosynthetic process (Table S1) consistent with the expectation that these are non-essential genes expressed in optimal conditions that are shut-down under thermal stress.

```{r exp_plot, echo=FALSE, warning = FALSE, fig.cap = "Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. carolinensis* (red) species"}

ggplot(Ap.df, aes(x=max.val, fill=colony)) +
  geom_density(alpha = 0.2, position = "identity") +
  scale_y_continuous(name = "Density") +
  scale_x_continuous(name = "Temperature of maximum expression")

```

As these genes represent only a handful of the responsive genes, the expression patterns of all responsive genes can be searched and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].

In addition, we examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses....

@@

We quantified differences in the reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$). Overall, we found that there was no significant difference in position of the mean (*t* = -0.95, *P* = 0.34), but changes in the shape due to slope (*t* = 24.67, *P* < 0.001), curvature (*t* = 15.71, *P* < 0.001) and wiggle (*t* = -12.45, *P* < 0.001) were all highly significant. On average, slope (95% CI 0.41 - 0.48) and curvature (95% CI 0.25 - 0.33) were greater in *A. picea* than *A. carolinensis*, indicating a steeper response to temperature.

@@

```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for genes that have increased expression at *Low* and *High* temperatures. Bars are Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. carolinensis* (red) species"}
T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
    geom_histogram(position = "dodge", binwidth = 3) +
    geom_density(adjust = 3, fill = NA, size = 2) + 
    scale_y_continuous(name="Density") +
    scale_x_continuous(name=expression("Temperature"))
print(T_on_plot)
```

>>>>

While characterizing the reactionome is useful for description of the molecular pathways for thermal tolerance, a greater understanding of the process of evolution can be gained using a comparative approach [@harvey1991]. Specifically, comparisons of populations or species in contrasting environments can reveal the extent to which geographic differentiation and selection has changed the number of genes involved in thermal tolerance. To accomplish this, we compared the thermal reactionomes between the two species of *Aphaenogaster*. *A. picea*, the northern species, experiences less chronic and acute heat stress than *A. carolensis*. Thus, we predicted that *A. picea* would have more genes than *A. carolinensis* that are up-regulated to high temperatures, while *A. carolinensis* would have higher expression of these thermally-tolerant genes near optimal temperatures. Reciprocally, we expected *A. carolinensis* to have greater up-regulation of genes in response to cold temperatures, while *A. picea* would have greater basal expression of cold-tolerant genes. Finally, phenotypic studies in *Aphaenogaster* [@warren2013] and other ecotherms [@hoffmann2013] have found less variation in critical maximum than minimum temperatures. From this, we hypothesized that if the species had more similar maximum temperatures than minimum temperatures, the northern species *A. picea* would have a broader thermal breadth of gene expression than *A. carolinensis*.

@@

To gain greater insight on the ecological genetics of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two species focusing on the thermally-responsive genes that had a significant species x temperature interaction. For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, *High*, *Low*, *Intermediate*, and *Bimodal*, as well as a fifth category *Not Responsive*. To focus on differences in gene expression between the two species, we performed GSEA for the set difference of each category.

@@

To test if genes that were responsive to cold temperatures in the southern *A. carolinensis* species or hot temperatures in the northern *A. picea* species had higher basal expression in the contrasting species, we performed non-parametric two-sided Wilcoxon tests comparing expression values near the optimum of 19.5° C.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
options(scipen = 1, digits = 3)
load("../ms.RData")
library(data.table)
library(pander)
library(ggplot2)
library(topGO)
```

The Trinity *de novo* transcriptome assembly included 126,172 genes with a total length of 100 million bp. Filtering to remove redundant or chimeric reads resulted in an assembly with 105,536 genes. The total length was 63 million bp with an N~50~ length of 895 bp and a mean gene size of 593 bp. Of the 105,536 filtered genes, 55,432 had hits to the NCBI-nr database. Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage. DeconSeq identified 5,675 genes as known contaminants from human, virus, bacterial or archael sources, leaving 99,861 clean transcripts. Of these, 47% have best hits to genes from the class Insecta [(Supplementary Krona plot)](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).

@@

We identified `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` genes with overall model fits of *P* < 0.05, retaining `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` after adjusting *P* values for false discovery rate. Of these genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` significantly change expression with temperature including `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` that have a species x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not thermally-responsive in one species (*NotResp*) differed significantly between the species ($\chi^2$ = `r chi1$statistic`, P < `r max(chi1$p.value, 0.001)`; Table 1). For both species, the majority of responsive transcripts were in the *Low* category (Fig. 1). More transcripts had *Intermediate* expression in *A. carolinensis* than in *A. picea*, which had nearly twice as many bimodally-expressed transcripts  (Fig. 1, Table 1).

```{r expression_type_table, echo=FALSE, results='asis'}
pandoc.table(tt3, style="rmarkdown", split.tables = Inf, caption = "Number of thermally-responsive genes in each expression category for each species. *AcNC* in rows and *ApVT* in columns.")
```

@@

```{r expression_type_plot, echo=FALSE, fig.cap = "Mosaic plot that shows the proportion of thermally-responsive genes in each expression category for each species."}
mosaicplot(t(tt2), shade = FALSE, ylab = "AcNC", xlab = "ApVT", main = "", cex = 0.5)
```

As the molecular pathways involved in thermal response to cold and hot temperatures differ, we performed gene set enrichment analysis on each expression category separately. Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "High"), ])` enriched (*P* < 0.01) for *High* transcripts, including categories such as `r Ap.gsea1[which(Ap.gsea1$Type == "High"), "Term"][1]` and `r Ap.gsea1[which(Ap.gsea1$Type == "High"), "Term"][6]` (Table S1). For *Low* genes, `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Low"), ])` GO terms were enriched including some implicated in stress response (e.g. `r Ap.gsea1[which(Ap.gsea1$Type == "Low"), "Term"][18]`) and cold tolerance (e.g. `r Ap.gsea1[which(Ap.gsea1$Type == "Low"), "Term"][6]`) (Table S1). For *Bimodal* genes, `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Bimodal"), ])` GO terms were enriched including `r length(grep("regulation", Ap.gsea1[which(Ap.gsea1$Type == "Bimodal"), "Term"]))` *regulation* terms (e.g. regulation of gene expression) indicating that these are general genes involved in activating responses to change (Table S1). Finally, for *Intermediate* genes `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Intermediate"), ])` GO terms were enriched, primarily metabolic or biosynthetic processes (Table S1).

In addition, we examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Figure 2). Though we found `r resp.stress.responsive.count` genes annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r GO.stress.Xsq$statistic`, P < `r round(GO.stress.Xsq$p.value, 2)`).

```{r rxn_plot, echo = FALSE, warning = FALSE, fig.cap = "Gene expression in transcripts per million (TPM) for thermally-reactive genes in the GO category *response to stress* for each colony"}
rxnplot <- ggplot(resp.TPM.dt.sub[GO0006950.responsive$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) +
  facet_grid(. ~ colony2) +
  scale_y_continuous(name = "Expression (scaled)") +
  scale_x_continuous(name = expression(paste("Temperature", degree, "C")))
print(rxnplot)
```

As these genes represent only a handful of the responsive genes, the expression patterns of all responsive genes can be searched by functional annotation or GO category and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].

@@

We quantified differences in the expression reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$). Overall, we found mean expression was greater in *AcNC* (*t*~`r t.offset$parameter`~ = `r t.offset$statistic`, *P* < `r max(t.offset$p.value, 0.001)`),

but changes in the shape due to slope , curvature (*t* = 15.71, *P* < 0.001) and wiggle (*t* = -12.45, *P* < 0.001) were all highly significant. On average, slope (95% CI 0.41 - 0.48) and curvature (95% CI 0.25 - 0.33) were greater in *A. picea* than *A. carolinensis*, indicating a steeper response to temperature.

```{r exp_plot, echo=FALSE, warning = FALSE, fig.cap = "Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. carolinensis* (red) species"}
ggplot(Ap.df, aes(x=max.val, fill=colony, colour = colony)) +
  geom_density(alpha = 0.2, position = "identity") +
  scale_y_continuous(name = "Density") +
  scale_x_continuous(name = "Temperature of maximum expression")
```

@@

```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for genes that have increased expression at *Low* and *High* temperatures. Bars are Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. carolinensis* (red) species"}
T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
    geom_histogram(position = "dodge", binwidth = 3) +
    geom_density(alpha = 0.2, adjust = 3) + 
    scale_y_continuous(name="Density") +
    scale_x_continuous(name=expression("Temperature"))
print(T_on_plot)
```
